### 2021-11-16 documents/sections/verbs.tex
discussion and factor comparison
'to be'

\section{Resistant verbs from a comparative perspective}
\label{sec:verbs}
In \cref{sec:extensions}, we introduced six distinct extensions of personal prefixes into \gl{1}\gl{s_a_} territory, and identified verbs resistant to each innovation.
The set of unaffected verbs was rather small in most cases, and (proto-)languages show a considerable degree of overlap in what verbs are unaffected.
In this section, we present these verbs from a comparative perspective.
\cref{sec:be} treats both roots of the copula \rc{eti}/\obj{a[p]} \qu{to be}, \cref{sec:say} \rc{ka[ti]} \qu{to say}, \cref{sec:go} \rc{ɨtə[mə]} \qu{to go}, and \cref{sec:come} \rc{(ət)jəpɨ} \qu{to come}; these are all verbs that \textcite{gildea2007greenberg} reconstruct as \gl{s_a_} verbs that were not derived from transitive verbs.
\cref{sec:godown} takes a look at \rc{ɨpɨtə} \qu{to go down}, which is resistant in \PTir and \PPek, and  \cref{sec:bathe} investigates \rc{e-pɨ} \qu{to bathe}, of which the \PPek reflex \rc{i-pɨ} resisted the extension of \rc{k-}.
The \obj{e}-initial verbs not affected by the extension of \rc{k-} in \akuriyo \pcref{sec:akuriyo} will not be discussed here, as they are a large and phonologically coherent group.

\subsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
For a comprehensive comparative overview for this verb, we refer the reader to \textcite[375--382]{gildea2018reconstructing}, who reconstructs two distinct roots serving as verbs \qu{to be} in modern Cariban languages.
One is the older copula \rc{a[p]}, which can be reconstructed as already having various irregularities in \PC.
The other is a root \rc{eti} reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Various modern languages use reflexes of these two roots in a suppletive manner, conditioned by person and\slash{}or \gl{tam} value.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while \obj{eʔi} is not attested in a first-person form. 
\carijo innovated \obj{j-}, but only in the \obj{et͡ʃi} root allomorph \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa innovated \obj{j-} for reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

%\subsection{\trio \obj{weka}/\obj{oeka} \qu{to defecate}}
%\label{sec:shit}
%The comparative picture for the resistant \trio \gl{s_a_} verb \qu{to defecate} \pcref{tab:defecate} shows that \rc{weka} was clearly originally an \gl{s_p_} verb.
%As in the case of \qu{to go down} \pcref{sec:godown}, this verb has a mixed \gl{s_a_}/\gl{s_p_} status in \wayana.
%The \bakairi reflex \obj{əeke} seems to contain a reflex of \rc{ət-}, which would explain its status as an \gl{s_a_} verb.
%While the class membership of \panare \obj{aiʔka} and \obj{iʔka} \qu{to defecate} is unknown, the former seems to contain a reflex of the detransitivizer, while the latter only consists of the root, meaning that one would assume \gl{s_a_} and \gl{s_p_} membership, respectively.
%
%\input{floats/defecate}
%
%\subsection{The \akuriyo movement verbs}
%\label{sec:movement}
%As discussed in \cref{sec:akuriyo}, four \akuriyo \gl{s_a_} verbs are attested as not having the first person marker \obj{t͡ʃ-} in \textcite{gildea1994akuriyo}; instead, they have \obj{w-} or its phonologically reduced form ∅.
%These verbs share two properties: they are all \obj{e-}initial, and they are all movement verbs.
%However, the same dataset also contains \obj{e-}initial movement verbs with a first person marker \obj{t͡ʃ-} \exref{aku-106}.
%As for the etymology of these four verbs, we can only speak to \obj{erama} \qu{to return}, which is a detransitivized form of \obj{rama} \qu{to put back}, a pair which is also found in other Cariban languages.
%For the other three verbs, we have not encountered potential cognates, but given their \obj{e-}initial nature and their semantics, it seems plausible that they, too, are derived from transitive verbs with meanings like \qu{to put up} etc.
%In any case, \akuriyo is only sparsely documented, and there are many open questions about it; the \akuriyo data in \textcite{gildea1994akuriyo} are not fully reliable.
%
%\ex<aku-106>\akuriyo \parencite[][84]{gildea1994akuriyo}\\
%\begingl
%\gla t͡ʃ-ewai//
%\glb \gl{1}-sit.down//
%\glft \qu{I sat down.}//
%\endgl
%\xe

>>>>

\section{Resistant verbs from a comparative perspective}
\label{sec:verbs}
In \cref{sec:extensions}, we introduced six distinct extensions of personal prefixes into \gl{1}\gl{s_a_} territory, and identified verbs resistant to each innovation.
The set of unaffected verbs is rather small in most cases, and (proto-)languages show a considerable degree of overlap in what verbs are unaffected.
Therefore, this section presents these verbs from a comparative perspective.
\cref{sec:be} treats both roots of the copula \rc{eti}/\obj{a[p]} \qu{to be}, \cref{sec:say} \rc{ka[ti]} \qu{to say}, \cref{sec:go} \rc{ɨtə[mə]} \qu{to go}, and \cref{sec:come} \rc{(ət)jəpɨ} \qu{to come}.
\cref{sec:godown} takes a look at \rc{ɨpɨtə} \qu{to go down}, which is resistant in \PTir and \PPek, and  \cref{sec:bathe} investigates \rc{e-pɨ} \qu{to bathe}, of which the \PPek reflex \rc{i-pɨ} resisted the extension of \rc{k-}.
The \obj{e}-initial verbs not affected by the extension of \obj{k-} in \akuriyo \pcref{sec:akuriyo} will not be discussed here, as they are a large and phonologically coherent group.

\subsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
For a comprehensive comparative overview for these two roots, we refer the reader to \textcite[375--382]{gildea2018reconstructing}.
\rc{a[p]} is the original copula and can be reconstructed as already having various irregularities in \PC.
\rc{eti}, the other root, is reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Various modern languages use reflexes of these two roots in a suppletive manner, conditioned by person and\slash{}or \gl{tam} value.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while \obj{eʔi} is not attested in a first-person form. 
\carijo innovated \obj{j-}, but only in the reflex of \rc{eti} \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa introduced \obj{j-} to the reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

%\subsection{\trio \obj{weka}/\obj{oeka} \qu{to defecate}}
%\label{sec:shit}
%The comparative picture for the resistant \trio \gl{s_a_} verb \qu{to defecate} \pcref{tab:defecate} shows that \rc{weka} was clearly originally an \gl{s_p_} verb.
%As in the case of \qu{to go down} \pcref{sec:godown}, this verb has a mixed \gl{s_a_}/\gl{s_p_} status in \wayana.
%The \bakairi reflex \obj{əeke} seems to contain a reflex of \rc{ət-}, which would explain its status as an \gl{s_a_} verb.
%While the class membership of \panare \obj{aiʔka} and \obj{iʔka} \qu{to defecate} is unknown, the former seems to contain a reflex of the detransitivizer, while the latter only consists of the root, meaning that one would assume \gl{s_a_} and \gl{s_p_} membership, respectively.
%
%\input{floats/defecate}
%
%\subsection{The \akuriyo movement verbs}
%\label{sec:movement}
%As discussed in \cref{sec:akuriyo}, four \akuriyo \gl{s_a_} verbs are attested as not having the first person marker \obj{t͡ʃ-} in \textcite{gildea1994akuriyo}; instead, they have \obj{w-} or its phonologically reduced form ∅.
%These verbs share two properties: they are all \obj{e-}initial, and they are all movement verbs.
%However, the same dataset also contains \obj{e-}initial movement verbs with a first person marker \obj{t͡ʃ-} \exref{aku-106}.
%As for the etymology of these four verbs, we can only speak to \obj{erama} \qu{to return}, which is a detransitivized form of \obj{rama} \qu{to put back}, a pair which is also found in other Cariban languages.
%For the other three verbs, we have not encountered potential cognates, but given their \obj{e-}initial nature and their semantics, it seems plausible that they, too, are derived from transitive verbs with meanings like \qu{to put up} etc.
%In any case, \akuriyo is only sparsely documented, and there are many open questions about it; the \akuriyo data in \textcite{gildea1994akuriyo} are not fully reliable.
